---
title: "airparif"
author: "Sarah Hosking"
date: "December 10, 2016"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(GGally)
library(lubridate)
library(reshape2) #plot missing vals

theme_set(theme_bw())

```

## R Markdown


```{r air pollution in paris Dec 2016}
airparif <- read.csv("airparif_centre_8_12_2016.csv", sep = ";", stringsAsFactors = FALSE)
#airparif <- read.csv("airparif_centre_8_12_2016.csv", sep = ";")

dim(airparif)
str(airparif)
summary(airparif)
head(airparif,20)
tail(airparif,20)

```

There seems to be a LOT of NA values

```{r}

ggplot_missing <- function(x){
  
  x %>% 
    is.na %>%
    melt %>%
    ggplot(data = .,
           aes(x = Var2,
               y = Var1)) +
    geom_raster(aes(fill = value)) +
    scale_fill_grey(name = "",
                    labels = c("Present","Missing")) +
    theme_minimal() + 
    theme(axis.text.x  = element_text(angle=45, vjust=0.5)) + 
    labs(x = "Variables in Dataset",
         y = "Rows / observations")
}

ggplot_missing(airparif[,1:7])
```


## Prepare the data

Where is the NA date?

```{r}

which(is.na(airparif$date))

# look at vals before & after this row
airparif[211:214,]

# okay this happened on '29/07/2011' @ 20h
airparif$date[213] <- '29/07/2011'
airparif[211:214,]

```


Let's convert the date var into a true Date type variable

```{r convert str date into date}

airparif$date <- as.Date(airparif$date, format = "%d/%m/%Y") 
str(airparif)

```

Then we can use tidyR's `separate` function to split these into day, month and year variables.

```{r split date into y m d vars}

# remove 1st row
airparif <- airparif[-1,]

# make a copy
airparif$date2 <- airparif$date

airparif <- separate(airparif, date2, c("year", "month", "day"), sep="-")
str(airparif)

#fix malformed date for row 212
#airparif$date[212] <- "2011-07-20"

#try again
#airparif <- separate(airparif, date, c("year", "month", "day"), sep="-")

```

Time to look at the distributions

First let's understand what these pollutants are

```{r}

# get column names
names(airparif)


```


What the abbreviations mean:

* `PM25` = fine particulate > 2.5 mm
* `PM10` = fine particulate > 10 mm
* `03` = Ozone
* `NO2` = Nitrogen Dioxide (Azote)
* `CO` = Carbon monoxide

Ok, so let's look at the data!

```{r plot histograms}

qplot(data = airparif, PM10)

```

Ugh. What's wrong here?

```{r}

str(airparif)

```

```{r convert columns to numeric}

# create list of cols to convert
cols<- names(airparif)
cols.num <- cols[3:length(airparif)]

# convert pollutants to numeric
airparif[cols.num] <- lapply(airparif[cols.num], as.numeric)


# this converts everything to a list
# need to reconvert to df
airparif <- data.frame(airparif)
str(airparif)
summary(airparif)


# plot again
qplot(data = airparif, PM10)

# increase # of bins
qplot(data = airparif, PM10, binwidth = 5)


```

Now let's try faceting

```{r facets}
# by year
qplot(data = airparif, PM10, binwidth = 5, facets = ~year)

# remove the NA values
qplot(data = na.omit(airparif), PM10, binwidth = 5, facets = ~year)

# what were the max values for each year?
tapply(airparif$PM10, airparif$year, summary)

ggplot(aes(PM10), data = na.omit(airparif)) +
  geom_histogram(binwidth = 5) +
  facet_wrap(~year, ncol = 1)

```


Looking at the other variables

```{r}

qplot(data = na.omit(airparif), PM25, binwidth = 5, facets = ~heure)
qplot(data = airparif, NO2, binwidth = 7, facets = ~heure)
qplot(data = airparif, O3, binwidth = 3, facets = ~year)
qplot(data = na.omit(airparif), CO, facets = ~year)


```

```{r boxplots}

ggplot(aes(x = as.factor(year), y = PM10), data = na.omit(airparif)) +
  geom_boxplot(na.rm = TRUE)

ggplot(aes(x = as.factor(year), y = PM25), data = airparif) +
  geom_boxplot()

ggplot(aes(x = as.factor(year), y = NO2), data = airparif) +
  geom_boxplot()

ggplot(aes(x = as.factor(year), y = O3), data = airparif) +
  geom_boxplot()

```

## Scatterplots

How has pollution evolved over time?

```{r}

ggplot(data = airparif, aes(x = date, y = PM25)) +
  geom_point(alpha = 0.2)

```




## Comparing variables

```{r, eval=FALSE, include=FALSE}

# create a sample
set.seed(888)
sample <- subset(airparif, select = c(PM10, PM25, NO2, O3, CO))
sample <- sample[sample(1:nrow(sample),2500),]

# look at correlations
ggpairs(sample, axisLabels = 'internal',
        lower = list(continuous = wrap("smooth", alpha=0.2, shape = I('.'))), 
        upper = list(combo = wrap("box", outlier.shape = I('.'))))
```

## Create summaries

```{r}
  
summary <- airparif %>% 
  group_by(year, date) %>% 
  summarise_each(funs(mean), PM10, PM25, NO2, O3, CO) %>% 
  ungroup() 

#mutate(yr.month = year * 100 + month) %>% 
  #group_by(yr.month) %>% 
  #arrange(yr.month)'''
  

head(summary)
tail(summary)

```

```{r}

# plot trend

ggplot(aes(date, PM10), data = na.omit(summary)) +
  geom_line()
  

```


Need to add some more context to this plot. What are the levels for medium, high, and very high pollution?

```{r adding h lines}

ggplot(aes(date, PM10), data = summary) +
  geom_line() +
  geom_hline(yintercept = 25, col = "green") +
  geom_hline(yintercept = 50, col = "yellow") +
  geom_hline(yintercept = 75, col = "orange") +
  geom_hline(yintercept = 100, col = "red")
```

So how many days have their been for each level?

```{r gather & spread}

# add a levels column to summary

#x <- list(1:3, 4:6)
#do.call(cbind, lapply(x, function(i) rowMeans(dat[, i])))

x = summary[2:7]
x

# create function to set pollution levels
pollution.level <- function(x, lower = 0, upper = 100, 
                            by = 25, sep = "-", above.char = "+") {
 labs <- c("very low", "low", "medium", "high", "very high")

 cut(floor(x), 
     breaks = c(seq(lower, upper, by = by), Inf),
     right = FALSE, 
     labels = labs)
}

# apply to all the columns of x
x[2:6] <- lapply(x[2:6], pollution.level)

```

```{r}

levels <- cbind(summary[1], x)

# gather(cases, "year", "n", 2:4)
levels.long <- gather(levels, "pollutant", "level", 3:7)

#spread(pollution, size, amount)

# reshaping to long format
count.levels.l <- levels.long %>% 
  # collapse pollutant columns into one
  # create year column
  # mutate('year' = year(date)) %>% 
  # create new 'level' column
  #gather(pollutant, level, 2:7) %>% 
  group_by(year,pollutant, level) %>% 
  # count number of days at each level
  summarise(n = n())

str(count.levels.l)

# need to conver 'level' col to ordered factor
count.levels.l$level <- ordered(count.levels.l$level, 
                                levels = c("very low", "low", "medium", 
                                           "high", "very high", "NA"))

# now reshape to wide format, count days, 
# fill missing vals with zeros
count.levels.w <- count.levels.l %>%
  spread(level, n, fill=0)

# and let's gather once more
test <- count.levels.w %>%
  ungroup() %>%
  group_by(year) %>%
  gather(year, pollutant, level, 3:8)
  


```

Ok, finally. Time to plot. 

```{r}

ggplot(data = filter(count.levels.l, year != 2011), aes(x=level, y=n,
                                  group = pollutant, 
                                  fill = pollutant)) +
  geom_bar(stat="identity") + 
  scale_x_discrete("Level") +
  scale_y_continuous("Number of Days") +
  facet_wrap(~pollutant, ncol=2) +
  #facet_wrap(~year)
```

Hmmmm...the problem is, the counts don't make sense, particularly as we haven't faceted by year.

Time to look at a breakdown

```{r}


```

